\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks

### 2016-01-12 nnt_manu.Rnw
no comments in code chunks

The stable release of \pkg{NeuralNetTools} can be installed from \ac{CRAN} and the development version can be installed from GitHub:

<<eval = F, echo = T>>=
# install from CRAN
install.packages('NeuralNetTools')
library(NeuralNetTools)

# install from GitHub
install.package('devtools')
library(devtools)
install_github('fawda123/NeuralNetTools')
library(NeuralNetTools)
@

@@

<<eval = T, echo = F, message = F>>=
library(NeuralNetTools)
@
<<results = hide>>=
# set random seed for initial weights for training
set.seed(123)

# mlp object, RSNNS package
library(RSNNS)
x <- neuraldat[, c('X1', 'X2', 'X3')]
y <- neuraldat[, 'Y1']
mod1 <- mlp(x, y, size = 5)

# nn object, neuralnet package
library(neuralnet)
mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)

#nnet object, nnet package
library(nnet)
mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
@

@@

# plot, pruned connections are omitted (default) or included with prune_col
pdf('figs/plotprune.pdf', height = 7, width = 7.7, family = 'serif')
par(mar = c(0, 0, 0, 0))
plotnet(mod, rel_rsc = c(3, 8))
plotnet(mod, prune_col = 'lightblue', rel_rsc = c(3, 8))
dev.off()
@
<<eval = F, echo = T, results = hide>>=
# pruned model using RSNNS
pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
  no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
  init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
 pruneFuncParams = pruneFuncParams)

# plot, pruned connections are omitted (default) or included with prune_col
par(mar = c(0, 0, 0, 0))
plotnet(mod, rel_rsc = c(3, 8))
plotnet(mod, prune_col = 'lightblue', rel_rsc = c(3, 8))
@
\begin{figure}[!ht]
\subfloat[][]{
\includegraphics[width=0.48\textwidth, page=1]{figs/plotprune.pdf}
\label{fig:plotprune1}
}
\subfloat[][]{
\includegraphics[width=0.48\textwidth, page=2]{figs/plotprune.pdf}
\label{fig:plotprune2}
}
\caption{A pruned neural network from \pkg{RSNNS} \citep{Bergmeir12} using the `Optimal Brain Surgeon' algorithm described in \citet{Zell98}.  The default plotting behavior of \code{plotnet} is to omit pruned connections (\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \code{prune\_col} argument (\ref{fig:plotprune2}).}
\label{fig:plotprune}
\end{figure}

@@

Both functions have similar implementations and require only a model object as input.  The default output is a \pkg{ggplot2} bar plot \citep[i.e., \code{geom\_bar},][]{Wickham09} that shows the relative importance of each input variable in the model (\cref{fig:plotimp}).  The plot aesthetics are based on internal code that can be changed using conventional syntax for \pkg{ggplot2} applied to the output object.  The importance values can also be returned as a \code{data.frame} if \code{bar\_plot = FALSE}.

<<eval = F>>=
# garson and olden functions for each model
garson(mod1)
olden(mod1)
garson(mod2)
olden(mod2)
garson(mod3)
olden(mod3)
@
<<results = hide, echo = FALSE>>=
pdf('figs/plotimp.pdf', family = 'serif', height = 2, width = 4)
garson(mod1)
olden(mod1)
garson(mod2)
olden(mod2)
garson(mod3)
olden(mod3)
dev.off()
@
\begin{figure}[!h]
\centering
\subfloat[][Model 1 (\code{mlp}), \code{garson}]{
\includegraphics[width = 0.47\textwidth, page = 1]{figs/plotimp.pdf}
\label{fig:plotimp1}
}
\subfloat[][Model 1 (\code{mlp}), \code{olden}]{
\includegraphics[width = 0.47\textwidth, page = 2]{figs/plotimp.pdf}
\label{fig:plotimp2}
}

@@

A second implementation of \code{lekprofile} is to group the unevaluated explanatory variables by natural groupings defined by the data. Covariance among predictors may present unlikely scenarios if all unevaluated variables are held at the same level (e.g., high values for one variable may be unlikely with high values for a second variable). The second option holds unevaluated variables at mean values defined by natural clusters in the data (\cref{fig:plotlek_sens2,fig:plotlek_bars2}). Clusters are identified using {\it k}-means clustering (\code{kmeans} from base \pkg{stats}, \citealt{Hartigan79}) of the input variables if the argument passed to \code{group_vals} is an integer value greater than one. The centers (means) of the clusters are then used as constant values for the unevaluated variables.  \citet{Beck14a} provide an example of the clustering method for \code{lekprofile} by evaluating response of a lake health index to different explanatory variables.  Lake clusters were identified given covariance among variables, such that holding explanatory variables at values defined by clusters created more ecologically valid response curves.  Both methods return similar plots, with additional options to visualize the groupings for unevaluated explanatory variables (\cref{fig:plotlek_bars}).  For the latter case, \code{split\_show = TRUE} will return a stacked bar plot for each group with heights within each bar proportional to the constant values.
<<eval = F>>=
# lek profile, quantile evaluation
lekprofile(mod3)

# lek profile, quantile groups 
lekprofile(mod3, group_show = TRUE)

# lek profile, cluster evaluation
lekprofile(mod3, group_vals = 6)

# lek profile, cluster groups
lekprofile(mod3, group_vals = 6, group_show = TRUE)
@
<<echo = F, results = hide>>=
pdf('figs/plotlek_sens.pdf', height = 2.5, width = 8, family = 'serif')
lekprofile(mod3) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
lekprofile(mod3, group_vals = 6) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
dev.off()

>>>>

The stable release of \pkg{NeuralNetTools} can be installed from \ac{CRAN} and loaded as follows:

<<eval = F, echo = T>>=
install.packages('NeuralNetTools')
library(NeuralNetTools)
@

@@

<<eval = T, echo = F, message = F>>=
library(NeuralNetTools)
@
<<results = hide>>=
set.seed(123)
library(RSNNS)
x <- neuraldat[, c('X1', 'X2', 'X3')]
y <- neuraldat[, 'Y1']
mod1 <- mlp(x, y, size = 5)
library(neuralnet)
mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
library(nnet)
mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
@

@@

# plot, pruned connections are omitted (default) or included with prune_col
pdf('figs/plotprune.pdf', height = 7, width = 7.7, family = 'serif')
par(mar = c(0, 0, 0, 0))
plotnet(mod, rel_rsc = c(3, 8))
plotnet(mod, prune_col = 'lightblue', rel_rsc = c(3, 8))
dev.off()
@
<<eval = F, echo = T, results = hide>>=
pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
  no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
  init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
 pruneFuncParams = pruneFuncParams)
plotnet(mod, rel_rsc = c(3, 8))
plotnet(mod, prune_col = 'lightblue', rel_rsc = c(3, 8))
@
\begin{figure}[!ht]
\subfloat[][]{
\includegraphics[width=0.48\textwidth, page=1]{figs/plotprune.pdf}
\label{fig:plotprune1}
}
\subfloat[][]{
\includegraphics[width=0.48\textwidth, page=2]{figs/plotprune.pdf}
\label{fig:plotprune2}
}
\caption{A pruned neural network from \pkg{RSNNS} \citep{Bergmeir12} using the `Optimal Brain Surgeon' algorithm described in \citet{Zell98}.  The default plotting behavior of \code{plotnet} is to omit pruned connections (\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \code{prune\_col} argument (\ref{fig:plotprune2}).}
\label{fig:plotprune}
\end{figure}

@@

Both functions have similar implementations and require only a model object as input.  The default output is a \pkg{ggplot2} bar plot \citep[i.e., \code{geom\_bar},][]{Wickham09} that shows the relative importance of each input variable in the model (\cref{fig:plotimp}).  The plot aesthetics are based on internal code that can be changed using conventional syntax for \pkg{ggplot2} applied to the output object.  The importance values can also be returned as a \code{data.frame} if \code{bar\_plot = FALSE}.  Variable importance shown in \Cref{fig:plotimp} was estimated for each model:
<<eval = F>>=
garson(mod1)
olden(mod1)
garson(mod2)
olden(mod2)
garson(mod3)
olden(mod3)
@
<<results = hide, echo = FALSE>>=
pdf('figs/plotimp.pdf', family = 'serif', height = 2, width = 4)
garson(mod1)
olden(mod1)
garson(mod2)
olden(mod2)
garson(mod3)
olden(mod3)
dev.off()
@
\begin{figure}[!h]
\centering
\subfloat[][Model 1 (\code{mlp}), \code{garson}]{
\includegraphics[width = 0.47\textwidth, page = 1]{figs/plotimp.pdf}
\label{fig:plotimp1}
}
\subfloat[][Model 1 (\code{mlp}), \code{olden}]{
\includegraphics[width = 0.47\textwidth, page = 2]{figs/plotimp.pdf}
\label{fig:plotimp2}
}

@@

A second implementation of \code{lekprofile} is to group the unevaluated explanatory variables by natural groupings defined by the data. Covariance among predictors may present unlikely scenarios if all unevaluated variables are held at the same level (e.g., high values for one variable may be unlikely with high values for a second variable). The second option holds unevaluated variables at mean values defined by natural clusters in the data (\cref{fig:plotlek_sens2,fig:plotlek_bars2}). Clusters are identified using {\it k}-means clustering (\code{kmeans} from base \pkg{stats}, \citealt{Hartigan79}) of the input variables if the argument passed to \code{group_vals} is an integer value greater than one. The centers (means) of the clusters are then used as constant values for the unevaluated variables.  \citet{Beck14a} provide an example of the clustering method for \code{lekprofile} by evaluating response of a lake health index to different explanatory variables.  Lake clusters were identified given covariance among variables, such that holding explanatory variables at values defined by clusters created more ecologically valid response curves.  Both methods return similar plots, with additional options to visualize the groupings for unevaluated explanatory variables (\cref{fig:plotlek_bars}).  For the latter case, \code{group\_show = TRUE} will return a stacked bar plot for each group with heights within each bar proportional to the constant values.  Sensitivity profiles were created using the standard approach based on quantiles and using the alternative clustering method (\cref{fig:plotlek_sens}), including bar plots of the relative values for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
<<eval = F>>=
lekprofile(mod3)
lekprofile(mod3, group_show = TRUE)
lekprofile(mod3, group_vals = 6)
lekprofile(mod3, group_vals = 6, group_show = TRUE)
@
<<echo = F, results = hide>>=
pdf('figs/plotlek_sens.pdf', height = 2.5, width = 8, family = 'serif')
lekprofile(mod3) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
lekprofile(mod3, group_vals = 6) + theme(plot.margin = grid::unit(c(0, 0, 0, 0), 'cm'))
dev.off()

### 2016-01-12 nnt_manu.tex
no comments in code chunks

The stable release of \pkg{NeuralNetTools} can be installed from \ac{CRAN} and the development version can be installed from GitHub:

\begin{Schunk}
\begin{Sinput}
R> # install from CRAN
R> install.packages('NeuralNetTools')
R> library(NeuralNetTools)
R> # install from GitHub
R> install.package('devtools')
R> library(devtools)
R> install_github('fawda123/NeuralNetTools')
R> library(NeuralNetTools)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> # set random seed for initial weights for training
R> set.seed(123)
R> # mlp object, RSNNS package
R> library(RSNNS)
R> x <- neuraldat[, c('X1', 'X2', 'X3')]
R> y <- neuraldat[, 'Y1']
R> mod1 <- mlp(x, y, size = 5)
R> # nn object, neuralnet package
R> library(neuralnet)
R> mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
R> #nnet object, nnet package
R> library(nnet)
R> mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> # pruned model using RSNNS
R> pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
+   no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
+   init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
R> mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
+  pruneFuncParams = pruneFuncParams)
R> # plot, pruned connections are omitted (default) or included with prune_col
R> par(mar = c(0, 0, 0, 0))
R> plotnet(mod, rel_rsc = c(3, 8))
R> plotnet(mod, prune_col = 'lightblue', rel_rsc = c(3, 8))
\end{Sinput}
\end{Schunk}
\begin{figure}[!ht]
\subfloat[][]{
\includegraphics[width=0.48\textwidth, page=1]{figs/plotprune.pdf}
\label{fig:plotprune1}
}
\subfloat[][]{
\includegraphics[width=0.48\textwidth, page=2]{figs/plotprune.pdf}
\label{fig:plotprune2}
}
\caption{A pruned neural network from \pkg{RSNNS} \citep{Bergmeir12} using the `Optimal Brain Surgeon' algorithm described in \citet{Zell98}.  The default plotting behavior of \code{plotnet} is to omit pruned connections (\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \code{prune\_col} argument (\ref{fig:plotprune2}).}
\label{fig:plotprune}
\end{figure}

@@

Both functions have similar implementations and require only a model object as input.  The default output is a \pkg{ggplot2} bar plot \citep[i.e., \code{geom\_bar},][]{Wickham09} that shows the relative importance of each input variable in the model (\cref{fig:plotimp}).  The plot aesthetics are based on internal code that can be changed using conventional syntax for \pkg{ggplot2} applied to the output object.  The importance values can also be returned as a \code{data.frame} if \code{bar\_plot = FALSE}.

\begin{Schunk}
\begin{Sinput}
R> # garson and olden functions for each model
R> garson(mod1)
R> olden(mod1)
R> garson(mod2)
R> olden(mod2)
R> garson(mod3)
R> olden(mod3)
\end{Sinput}
\end{Schunk}
\begin{figure}[!h]
\centering
\subfloat[][Model 1 (\code{mlp}), \code{garson}]{
\includegraphics[width = 0.47\textwidth, page = 1]{figs/plotimp.pdf}
\label{fig:plotimp1}
}
\subfloat[][Model 1 (\code{mlp}), \code{olden}]{
\includegraphics[width = 0.47\textwidth, page = 2]{figs/plotimp.pdf}
\label{fig:plotimp2}
}

@@

A second implementation of \code{lekprofile} is to group the unevaluated explanatory variables by natural groupings defined by the data. Covariance among predictors may present unlikely scenarios if all unevaluated variables are held at the same level (e.g., high values for one variable may be unlikely with high values for a second variable). The second option holds unevaluated variables at mean values defined by natural clusters in the data (\cref{fig:plotlek_sens2,fig:plotlek_bars2}). Clusters are identified using {\it k}-means clustering (\code{kmeans} from base \pkg{stats}, \citealt{Hartigan79}) of the input variables if the argument passed to \code{group_vals} is an integer value greater than one. The centers (means) of the clusters are then used as constant values for the unevaluated variables.  \citet{Beck14a} provide an example of the clustering method for \code{lekprofile} by evaluating response of a lake health index to different explanatory variables.  Lake clusters were identified given covariance among variables, such that holding explanatory variables at values defined by clusters created more ecologically valid response curves.  Both methods return similar plots, with additional options to visualize the groupings for unevaluated explanatory variables (\cref{fig:plotlek_bars}).  For the latter case, \code{split\_show = TRUE} will return a stacked bar plot for each group with heights within each bar proportional to the constant values.
\begin{Schunk}
\begin{Sinput}
R> # lek profile, quantile evaluation
R> lekprofile(mod3)
R> # lek profile, quantile groups 
R> lekprofile(mod3, group_show = TRUE)
R> # lek profile, cluster evaluation
R> lekprofile(mod3, group_vals = 6)
R> # lek profile, cluster groups
R> lekprofile(mod3, group_vals = 6, group_show = TRUE)
\end{Sinput}
\end{Schunk}
\begin{figure}[!ht]
\centering
\subfloat[][Response with quantiles]{
\includegraphics[width = 0.9\textwidth, page = 1]{figs/plotlek_sens.pdf}
\label{fig:plotlek_sens1}
}

>>>>

The stable release of \pkg{NeuralNetTools} can be installed from \ac{CRAN} and loaded as follows:

\begin{Schunk}
\begin{Sinput}
R> install.packages('NeuralNetTools')
R> library(NeuralNetTools)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> set.seed(123)
R> library(RSNNS)
R> x <- neuraldat[, c('X1', 'X2', 'X3')]
R> y <- neuraldat[, 'Y1']
R> mod1 <- mlp(x, y, size = 5)
R> library(neuralnet)
R> mod2 <- neuralnet(Y1 ~ X1 + X2 + X3, data = neuraldat, hidden = 5)
R> library(nnet)
R> mod3 <- nnet(Y1 ~ X1 + X2 + X3, data = neuraldat, size = 5)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> pruneFuncParams <- list(max_pr_error_increase = 10.0, pr_accepted_error = 1.0,
+   no_of_pr_retrain_cycles = 1000, min_error_to_stop = 0.01, 
+   init_matrix_value = 1e-6, input_pruning = TRUE, hidden_pruning = TRUE)
R> mod <- mlp(x, y, size = 5, pruneFunc = "OptimalBrainSurgeon",
+  pruneFuncParams = pruneFuncParams)
R> plotnet(mod, rel_rsc = c(3, 8))
R> plotnet(mod, prune_col = 'lightblue', rel_rsc = c(3, 8))
\end{Sinput}
\end{Schunk}
\begin{figure}[!ht]
\subfloat[][]{
\includegraphics[width=0.48\textwidth, page=1]{figs/plotprune.pdf}
\label{fig:plotprune1}
}
\subfloat[][]{
\includegraphics[width=0.48\textwidth, page=2]{figs/plotprune.pdf}
\label{fig:plotprune2}
}
\caption{A pruned neural network from \pkg{RSNNS} \citep{Bergmeir12} using the `Optimal Brain Surgeon' algorithm described in \citet{Zell98}.  The default plotting behavior of \code{plotnet} is to omit pruned connections (\ref{fig:plotprune1}), whereas they can be viewed as dashed lines by including the \code{prune\_col} argument (\ref{fig:plotprune2}).}
\label{fig:plotprune}
\end{figure}

@@

Both functions have similar implementations and require only a model object as input.  The default output is a \pkg{ggplot2} bar plot \citep[i.e., \code{geom\_bar},][]{Wickham09} that shows the relative importance of each input variable in the model (\cref{fig:plotimp}).  The plot aesthetics are based on internal code that can be changed using conventional syntax for \pkg{ggplot2} applied to the output object.  The importance values can also be returned as a \code{data.frame} if \code{bar\_plot = FALSE}.  Variable importance shown in \Cref{fig:plotimp} was estimated for each model:
\begin{Schunk}
\begin{Sinput}
R> garson(mod1)
R> olden(mod1)
R> garson(mod2)
R> olden(mod2)
R> garson(mod3)
R> olden(mod3)
\end{Sinput}
\end{Schunk}
\begin{figure}[!h]
\centering
\subfloat[][Model 1 (\code{mlp}), \code{garson}]{
\includegraphics[width = 0.47\textwidth, page = 1]{figs/plotimp.pdf}
\label{fig:plotimp1}
}
\subfloat[][Model 1 (\code{mlp}), \code{olden}]{
\includegraphics[width = 0.47\textwidth, page = 2]{figs/plotimp.pdf}
\label{fig:plotimp2}
}

@@

A second implementation of \code{lekprofile} is to group the unevaluated explanatory variables by natural groupings defined by the data. Covariance among predictors may present unlikely scenarios if all unevaluated variables are held at the same level (e.g., high values for one variable may be unlikely with high values for a second variable). The second option holds unevaluated variables at mean values defined by natural clusters in the data (\cref{fig:plotlek_sens2,fig:plotlek_bars2}). Clusters are identified using {\it k}-means clustering (\code{kmeans} from base \pkg{stats}, \citealt{Hartigan79}) of the input variables if the argument passed to \code{group_vals} is an integer value greater than one. The centers (means) of the clusters are then used as constant values for the unevaluated variables.  \citet{Beck14a} provide an example of the clustering method for \code{lekprofile} by evaluating response of a lake health index to different explanatory variables.  Lake clusters were identified given covariance among variables, such that holding explanatory variables at values defined by clusters created more ecologically valid response curves.  Both methods return similar plots, with additional options to visualize the groupings for unevaluated explanatory variables (\cref{fig:plotlek_bars}).  For the latter case, \code{group\_show = TRUE} will return a stacked bar plot for each group with heights within each bar proportional to the constant values.  Sensitivity profiles were created using the standard approach based on quantiles and using the alternative clustering method (\cref{fig:plotlek_sens}), including bar plots of the relative values for unevaluated explanatory variables (\cref{fig:plotlek_bars}).
\begin{Schunk}
\begin{Sinput}
R> lekprofile(mod3)
R> lekprofile(mod3, group_show = TRUE)
R> lekprofile(mod3, group_vals = 6)
R> lekprofile(mod3, group_vals = 6, group_show = TRUE)
\end{Sinput}
\end{Schunk}
\begin{figure}[!ht]
\centering
\subfloat[][Response with quantiles]{
\includegraphics[width = 0.9\textwidth, page = 1]{figs/plotlek_sens.pdf}
\label{fig:plotlek_sens1}
}
